WASP-4 is Accelerating Toward the Earth

### 2020-03-21 paper/ms.tex
updated mass/sma constraints in paper

\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be
  decreasing at a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.  Proposed
  explanations for the period change include tidal orbital decay,
  apsidal precession, and acceleration of the system along the line of
  sight.  To investigate further, we performed new radial velocity
  measurements and speckle imaging of WASP-4.  The radial-velocity
  data show that the system is accelerating towards the Sun at a rate
  of $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.  The
  associated Doppler effect should cause the apparent period to shrink
  at a rate of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.  Thus, the observed change in the transit
  period is mostly or entirely produced by the line-of-sight
  acceleration of the system.  This acceleration is probably caused by
  a wide-orbiting companion of mass 10-200$\,M_{\rm Jup}$ and orbital
  distance 10-100$\,$AU, based on the magnitude of the radial-velocity
  trend and the non-detection of any companion in the speckle images.
  We expect that the orbital periods of 1 out of 3 hot Jupiters will
  change at rates similar to WASP-4b, based on the hot-Jupiter
  companion statistics of \citet{knutson_friends_2014}.
  Continued radial velocity monitoring of hot Jupiters is therefore
  essential to distinguish the effects of tidal orbital decay or
  apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.8\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.8cm}
	\caption{
		{\bf Masses and semi-major axes of companions that meet
			requirements of both the radial velocities and the speckle
			imaging.} The likelihood inferred from radial velocities is shown
		in grayscale, and the region excluded from the speckle imaging is
		indicated.  The black line illustrates the expected degeneracy
		between mass and semimajor-axis.
		\label{fig:mass_sma}
		\vspace{-0cm}
	}
\end{figure*}

@@

First, we converted the contrast curves obtained from speckle imaging
(Figure~\ref{fig:zorro}) to upper limits on the mass of a companion as
a function of \added{projected} separation. To do this, we followed
\citet{montet_trends_2014}, and opted to employ the
\citet{baraffe_evolutionary_2003} models for substellar mass objects
and the MIST isochrones for stellar mass objects
\citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015,dotter_mesa_2016,choi_mesa_2016}.
We assumed that the system age was 5 Gyr, at which point the companion
would have fully contracted.

Due to the custom filters of the Zorro imager, and corresponding lack
of synthetic photometry, we further assumed that all sources had
blackbody spectra. While this is a simplification, we do not have
ready access to the planetary and stellar atmosphere models needed for
the consistent calculation with the \texttt{COND03} and \texttt{MESA}
models.  We adopted the effective temperatures and bolometric
luminosities from the \citet{baraffe_evolutionary_2003} and MIST
isochrones.  Using these theoretical quantities and the
empirically-measured Zorro bandpasses, we calculated absolute
magnitudes in the 562 and 832 nm Zorro bands for stellar and planetary
mass companions.  Applying the same calculation to WASP-4 itself using
the effective temperature and bolometric luminosity from
\citetalias{bouma_wasp4b_2019}, we derived the transformation from
contrast ratio to companion mass.  The resulting
limits \added{derived if we assume maximal projected separations} are
\replaced{shown}{shown with the blue line} in Figure~\ref{fig:mass_sma}.
\added{We describe our procedure for incorporating projection effects
in the following paragraphs.}

@@

\explain{Modified wording for clarity.}
\replaced{For each simulated companion, we drew a set of model
parameters from the converged posterior probability samples of our
model for the WASP-4 radial velocities.}{The orbital properties of the
inner hot Jupiter were assigned for each simulated system by sampling
from from the radial velocity posterior derived in
Section~\ref{subsec:rvs}.}  We subtracted the orbital component of the
model from the observed RVs, leaving behind the RV residuals with a
linear trend.  Given $(a_{\rm c}, M_{\rm c}, e_{\rm c})$ for each
simulated outer companion, and the choice of instrument offsets and
jitters, we performed a maximum likelihood fit for the time and
argument of periastron of the outer simulated companion.  We converted
the resulting $128\times128\times512$ cube of log-likelihood values to
probabilities, and averaged over the samples in each grid cell to
derive a probability distribution in mass and semi-major axis, based
on the radial velocities.

\added{We incorporated the speckle imaging limits in each simulated
system as follows. After fitting for the time and argument of
periastron, all the orbital parameters needed to find the projected
separation are known (we assumed uniform sensitivity as a function of
position angle, and therefore fixed the longitude of the
ascending node to zero). We calculated the projected separation using
the parametrization given by \citet{quirrenbach_2010}, which required
solving Kepler's equation for each simulated system at the time of the
imaging observation (September 28, 2019).   For each cell, we then
calculated the fraction of companions whose masses and
projected separations would have enabled detection.  }

\replaced{Figure~\ref{fig:mass_sma} shows the result.}
{The contours in Figure~\ref{fig:mass_sma} show the result of multiplying the radial
velocity and speckle imaging probability distributions. The companion
responsible for the acceleration has a true mass of 10-200$\,M_{\rm
Jup}$ and an orbital distance of $10$-$100\,{\rm AU}$.}
%FIXME: might need to update this text!

>>>>

\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be
  decreasing at a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.  Proposed
  explanations for the period change include tidal orbital decay,
  apsidal precession, and acceleration of the system along the line of
  sight.  To investigate further, we performed new radial velocity
  measurements and speckle imaging of WASP-4.  The radial-velocity
  data show that the system is accelerating towards the Sun at a rate
  of $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.  The
  associated Doppler effect should cause the apparent period to shrink
  at a rate of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.  Thus, the observed change in the transit
  period is mostly or entirely produced by the line-of-sight
  acceleration of the system.  This acceleration is probably caused by
  a wide-orbiting companion of mass 10-300$\,M_{\rm Jup}$ and orbital
  distance 10-100$\,$AU, based on the magnitude of the radial-velocity
  trend and the non-detection of any companion in the speckle images.
  We expect that the orbital periods of 1 out of 3 hot Jupiters will
  change at rates similar to WASP-4b, based on the hot-Jupiter
  companion statistics of \citet{knutson_friends_2014}.
  Continued radial velocity monitoring of hot Jupiters is therefore
  essential to distinguish the effects of tidal orbital decay or
  apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

\explain{We learned about this after submission.}
\subsection{Astrometry}

\added{
In review, it came to our attention that observations by the Gaia satellite...
%FIXME
}

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.8\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.8cm}
	\caption{
    {\bf Masses and semi-major axes of companions that meet
    requirements of both the radial velocities and the speckle
    imaging.} Contours show the joint radial velocity and speckle
    imaging probability density with 1, 2, and 3-$\sigma$
    significance. The dotted line shows the speckle imaging limits at
    maximal projected separation.  Relevant projection effects were
    marginalized out when calculating the contours.  The horizontal
    dashed line is the mass limit inferred from observations that
    WASP-4 is single-lined, requiring any companion to contribute no
    more than one-tenth of the observed light.  The black line shows
    the expected degeneracy between mass and semimajor-axis.
    \explain{Caption was modified to account for updated figure. The
    updated upper limit on the mass is now roughly 300$M_{\rm Jup}$,
    instead of 200$M_{\rm Jup}$. The text has been modified
    accordingly.} \label{fig:mass_sma}
		\vspace{-0cm}
	}
\end{figure*}

@@

First, we converted the contrast curves obtained from speckle imaging
(Figure~\ref{fig:zorro}) to upper limits on the companion mass as
a function of \added{projected} separation. To do this, we followed
\citet{montet_trends_2014}, and opted to employ the
\citet{baraffe_evolutionary_2003} models for substellar mass objects
and the MIST isochrones for stellar mass objects
\citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015,dotter_mesa_2016,choi_mesa_2016}.
We assumed that the system age was 5 Gyr, at which point the companion
would have fully contracted.

Due to the custom filters of the Zorro imager, and corresponding lack
of synthetic photometry, we further assumed that all sources had
blackbody spectra. While this is a simplification, we do not have
ready access to the planetary and stellar atmosphere models needed for
the consistent calculation with the \texttt{COND03} and \texttt{MESA}
models.  We adopted the effective temperatures and bolometric
luminosities from the \citet{baraffe_evolutionary_2003} and MIST
isochrones.  Using these theoretical quantities and the
empirically-measured Zorro bandpasses, we calculated absolute
magnitudes in the 562 and 832 nm Zorro bands for stellar and planetary
mass companions.  Applying the same calculation to WASP-4 itself using
the effective temperature and bolometric luminosity from
\citetalias{bouma_wasp4b_2019}, we derived the transformation from
contrast ratio to companion mass.  The resulting
limits \added{derived if we assume maximal projected separations} are
\replaced{shown}{shown with the dotted line} in Figure~\ref{fig:mass_sma}.
\added{
However, because we are observing the primary star accelerate towards
our line of sight, the companion very well be near
inferior conjunction.
We incorporate the relevant projection effects
as described in the following paragraphs.}

@@

\explain{Modified wording for clarity.}
\replaced{For each simulated companion, we drew a set of model
parameters from the converged posterior probability samples of our
model for the WASP-4 radial velocities.}{The orbital properties of the
inner hot Jupiter were assigned for each simulated system by sampling
from from the radial velocity posterior derived in
Section~\ref{subsec:rvs}.}  We subtracted the orbital component of the
model from the observed RVs, leaving behind the RV residuals with a
linear trend.  Given $(a_{\rm c}, M_{\rm c}, e_{\rm c})$ for each
simulated outer companion, and the choice of instrument offsets and
jitters, we performed a maximum likelihood fit for the time and
argument of periastron of the outer simulated companion.  \deleted{We converted
the resulting $128\times128\times512$ cube of log-likelihood values to
probabilities, and averaged over the samples in each grid cell to
derive a probability distribution in mass and semi-major axis, based
on the radial velocities.}

\added{We then incorporated the speckle imaging limits in each simulated
system as follows. After fitting for the time and argument of
periastron, all the orbital parameters needed to find the projected
separation are known (we assumed uniform sensitivity as a function of
position angle, and therefore fixed the longitude of the
ascending node to zero). We calculated the projected separation using
the parametrization given by \citet{quirrenbach_2010}, which required
solving Kepler's equation for each simulated system at the time of the
imaging observation (September 28, 2019).  If a simulated companion's
mass and projected separation put it above the 5$\sigma$ contrast curve,
we assumed it would have been detected.
We multiplied
the resulting $128\times128\times512$ cubes of radial velocity and speckle
imaging probabilities, and marginalized over the systems in each grid cell to
derive a probability distribution in mass and semi-major axis.
}
\replaced{Figure~\ref{fig:mass_sma} shows the result.}
{The contours in Figure~\ref{fig:mass_sma} show the result: the companion
responsible for the acceleration has a true mass of 10-300$\,M_{\rm
Jup}$ and an orbital distance of $10$-$100\,{\rm AU}$.}
%FIXME: might need to update this text!
